---
title: "PSTAT 126 Project"
author: Jeff Shen
date: "2023-04-27"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
# default code chunk options
knitr::opts_chunk$set(echo = T,
                      results = 'markup',
                      message = F, 
                      warning = F,
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center') 

# load packages
library(faraway)
library(tidyverse)
library(tidymodels)
library(modelr)
library(ggplot2)
library(glmnet)

```
## STEP TWO

### Introduction

In this project, we will analyze the relationship between income and education using the adult.csv dataset. Rather than simply comparing income levels and education levels, we will aggregate the data by calculating the percentage of individuals with income higher then 50k for each education level. This will allow us to create a quantitative variable for the proportion of individuals in each education level who earn a high income, which we can then use to explore the relationship between education and income. The dataset of interest comes from the UCI Machine Learning Repository and is available on Kaggle as well at: https://www.kaggle.com/datasets/uciml/adult-census-income. It contains information about individuals from the 1994 Census database, including demographic variables such as age, education, marital status, occupation, and more.

The variable of interest for this analysis is the income column, and we will be using a simple linear regression model to examine its relationship with education.

### Hypotheses


Our objective is to investigate the relationship between education level and income among individuals in the 1994 Census database by aggregating the data to calculate the percentage of individuals with income higher than 50k for each education level. We aim to determine whether there is a positive linear relationship between education level and the proportion of individuals with high income.

Null hypothesis: There is no significant positive linear relationship between education and the proportion of individuals with high income in the 1994 Census database. The beta coefficient for education (Î²1) in a linear regression model predicting the proportion of individuals with high income is equal to zero.

$$H_0: \beta_1 = 0$$

Alternative hypothesis: There is a significant positive linear relationship between education and the proportion of individuals with high income in the 1994 Census database, with higher levels of education associated with a higher proportion of individuals with high income. The beta coefficient for education (Î²1) in a linear regression model predicting the proportion of individuals with high income is greater than zero.

$$H_a: \beta_1 > 0$$

We can test these hypotheses using a linear regression model and examining the p-value associated with the beta coefficient for education. If the p-value is less than our chosen significance level (typically 0.05), we can reject the null hypothesis and conclude that there is evidence of a positive linear relationship between education and the proportion of individuals with high income.

### Checking the Assumptions for Linear Regression
```{r}
income_data <- read.csv("~/Desktop/adult.csv")
head(income_data)
```

### Plots and Transformations

```{r assumption_plots, fig.show = 'show', echo=FALSE}
# Divide into education levels
edu_levels <- c("Preschool", "1st-4th", "5th-6th", "7th-8th", "9th", "10th", "11th", "12th",
                "HS-grad", "Some-college", "Assoc-acdm", "Assoc-voc", "Bachelors", "Masters", "Prof-school", "Doctorate")

# Transform each education level to an 'education number' to compare 2 quantitative variables
income_data <- income_data %>%
  mutate(education_num = match(education, edu_levels))

# Calculate the percentage of individuals making over 50k in each education group to create data aggregation for income discrepancies 
income_summary <- income_data %>%
  group_by(education_num) %>%
  summarize(prop_over_50k = mean(income == ">50K"))
prop_over_50k <- income_summary$prop_over_50k
# Create a scatterplot of education level (numerical) vs. proportion over 50k
ggplot(income_summary, aes(x = education_num, y = prop_over_50k)) +
  geom_point() +
  labs(x = "Education Level", y = "Proportion Making Over 50K", title = "Scatter Plot") +
  scale_x_continuous(breaks = c(1:length(edu_levels)), labels = edu_levels) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title = element_text(size = 11, hjust = 0.5, vjust = 1.5))

# Create a bar graph of proportion of high income individuals by education level
ggplot(data=income_summary, aes(x = education_num, y = prop_over_50k)) +
  geom_bar(stat = "identity", fill = "#377EB8") +
  labs(x = "Education Level", y = "Proportion with Income over 50K", title = "Proportion of High Income by Education") +
  scale_x_continuous(breaks = c(1:length(edu_levels)), labels = edu_levels) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
    plot.title = element_text(size = 11, hjust = 0.5, vjust = 1.5))

# Create a stacked bar chart of income against education level
ggplot(data = income_data, aes(x = education_num, fill = income)) +
  geom_bar(position = "stack") +
  labs(x = "Education Level", y = "Count", title = "Distribution of Income by Education Level") +
  scale_x_continuous(breaks = 1:length(edu_levels), labels = edu_levels) +
  scale_fill_manual(values = c("#377EB8", "#E41A1C"), labels = c("<=50K", ">50K")) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title = element_text(size = 11, hjust = 0.5, vjust = 1.5))
```
Looking at the scatter plot above, it can be seen that the data roughly follows a linear path and has constant variability across the plot. There is no squeeze or diversion to indicate that there is different variabilities in the graph. However, we will still transform the graph in log and sqrt to see if there could be a better transformation that exists. 
```{r lm_plot, fig.show='show', echo=FALSE}
# Perform linear regression
lm_fit <- lm(prop_over_50k ~ education_num, data = income_summary)
summary(lm_fit)

# Plot trend line
ggplot(income_summary, aes(x = education_num, y = prop_over_50k)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Education Level", y = "Proportion Making Over 50K") +
  scale_x_continuous(breaks = c(1:length(edu_levels)), labels = edu_levels) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
    plot.title = element_text(size = 11, hjust = 0.5, vjust = 1.5))
```

```{r log_plot, fig.show='show', echo=FALSE}
# Generates a log transformation for the scatterplot
log_plot <- ggplot(income_summary, aes(x = log10(education_num), y = prop_over_50k)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Log Education Level", y = "Proportion Making Over 50K") +
  scale_x_continuous(breaks = log10(c(1:length(edu_levels))), labels = edu_levels) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title = element_text(size = 11, hjust = 0.5, vjust = 1.5))
log_plot
# Generates summary to view R^2 value for the log transformation
lm_log_fit <- lm(prop_over_50k ~ log10(education_num), data = income_summary)
summary(lm_log_fit)

```
```{r sqrt_plot, fig.show='show', echo=FALSE}
# Generates a sqrt transformation for the scatterplot
sqrt_plot <- ggplot(income_summary, aes(x = sqrt(education_num), y = prop_over_50k)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Square Root of Education Level", y = "Proportion Making Over 50K") +
  scale_x_continuous(breaks = sqrt(c(1:length(edu_levels))), labels = edu_levels) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title = element_text(size = 11, hjust = 0.5, vjust = 1.5))
sqrt_plot

# Generates summary to view R^2 value for the sqrt transformation
lm_sqrt_fit <- lm(prop_over_50k ~ sqrt(education_num), data = income_summary)
summary(lm_sqrt_fit)
```
#### Scatter Plot Analysis

Our initial scatter plot data compared education level (numerical) vs. proportion making over 50k. The graph displayed an exponential upward trend for the proportion making over 50k as the level of education increased. In order to find a better fit to the trend line, we created multiple transformations of the scatter plot in an attempt to make the plot more linear. For our transformations, we decided to manipulate the data by comparing both the log and square root of education level to the proportion making over 50k.

After analyzing the results from the new transformations, we determined that the log transformation plot created an even steeper rise in comparison to the original and as a result pushed the points even further away from the trend line. Although not as severe, the square root transformation also became steeper than the original. This meant that neither transformation linearized the data set and we concluded that the original plot showed the best fit. Furthermore, this could be confirmed by comparing the r-squared value for each of the three scatter plots. The original plot posted the highest value at 0.7872, compared to the log and square root plots at 0.4959 and 0.6595, respectively.

### Test of Beta 1 
```{r beta_coeff, echo=FALSE}
beta_1 <- coef(lm_fit)[2]
cat("beta_1 coefficient: ", beta_1, "\n")
```
Since $\beta_1 > 0$, we can reject the null hypothesis, and determine that there is in fact a positive linear relationship between education and income. 

### Confidence Interval
```{r CI_B1, echo=FALSE}

#confidence interval for education num
l.model <- lm(education_num ~ 1, income_summary)
confint(l.model, level = 0.95)

```

### Transformed Plot
```{r Transformed Plot, fig.show = 'show', echo=FALSE}
#plot of transformed variable for income data with income variable
p <- ggplot(income_summary, aes(x = education_num, y = prop_over_50k)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Education Level", y = "Income Proportion Over 50k")

p


#Confidence Interval for the mean and individual response of income



```

### Residuals Plots and R^2
```{r Residuals Plots and R^2, fig.show='show', echo = FALSE}
#Assess the fit of your model. Discuss the ð‘…2 value and the residual plot(s). Remember that residual plots (not ð‘…2) determine whether a linear model is appropriate

#calculating R^2 

l.model2 <- lm(prop_over_50k ~ education_num, income_summary)

rsquare(l.model2, income_summary)

#plotting the residuals

#calculating residuals 
res <- resid(l.model2)

#plot of fitted vs residual plots
plot(fitted(l.model2), res)

#adding line to the model 
abline(0,0)

#qq plot for residuals 
qqnorm(res)

#adding line to the plot of residuals 
qqline(res)

#density plot of residuals 
plot(density(res))


```


### Conclusion
